\bf Eye Fitting Straight Lines in the Modern Era

### 2022-01-24 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
working on submission

authors: 
- name: Emily A. Robinson 1
  #thanks: The authors gratefully acknowledge ...
  affiliation: Department of Statistics, University of Nebraska - Lincoln

@@

abstract: |
  Fitting lines by eye through a set of points has been explored since the 20th century. Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. In 2015, the New York Times introduced an interactive feature, called 'You Draw It'. Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality. The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics by drawing a line on their computer screen with their computer mouse. In my research, I established ‘You Draw It’ as a method for graphical testing by adapting the New York Times feature. I recruited participants via crowdsourcing websites and replicated the study found in Eye Fitting Straight Lines (Mosteller et al., 1981). Participants were directed to an RShiny application link and shown points following a linear trend and asked to draw a line through the data points using their computer mouse; task plots were generated using the r2d3 package in R statistical software. Results from my study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least-squares regression line. This trend was most prominent when shown data simulated with larger variances. The reproducibility of these results serves as evidence of the reliability of the you draw it method. Future work is necessary to implement the 'You Draw It' tool as a method of testing graphics. [200 word limit]

@@

# Introduction

Data visualization has become a central tool in modern data science.
@unwin2020data defines data visualization as the art of graphic displays to show data. 
Graphics illustrate the relationship between two sets of information or variables. 
Graphical displays are useful for data cleaning, exploratory data analysis, identifying trends, evaluating model output, and presenting results.
When creating a graphic, we must consider the design choices most effective for conveying the intended result.
In order to evaluate the effectiveness of a graphic, we must understand cognitive perception and biases related to different types of data structures. 
Efforts in the field of graphics have developed graphical testing tools and methods such as the lineup protocol [@buja2009statistical] to provide a framework for inferential testing.
Through experimentation, methods such as the lineup protocol allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation.

## Graph Perception

With the widespread use of graphics, work has been done to understand the effect of charts and graphs on human perception and higher cognitive processing such as learning and analysis [@green2009personal].
@carpenter1998model established the process of graph comprehension in which viewers interact with charts by first perceptually observing the visual features and later translating to cognitive processing of the information depicted by those features.
Much of the research on the perception of visual features in charts has been conducted in psychophysics and test for accuracy and quantitative comparisons when understanding a plot. 
Cleveland and McGill conducted a series of cognitive tasks designed to establish a hierarchy of visual components for making comparisons [@cleveland1984graphical]. For example, it is more effective to display information on an $x$ or $y$ axis rather than using color in order to reduce the visual effort necessary to make numerical comparisons.
Other early studies evaluated human perception of lines. 
For instance, @cleveland1985graphical found that assessing the position of points along an axis is easier than determining the slope of a line.
The results of these cognitive tasks provide some consistent guidance for chart design; however, other methods of visual testing can further evaluate design choices and help us understand cognitive biases related to the evaluation of statistical charts.

@@

Graphical tests are useful for studying the perception of statistical graphs. 
Studies might ask participants to identify differences in graphs, read information off of a chart accurately, use data to make correct real-world decisions, or predict the next few observations.
All of these types of tests require different levels of use and manipulation of the information being presented in the chart. Early researchers studied graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception]. 
These studies generally tested participants ability to detect a stimulus or a difference between two stimuli.
Here we focus on the how graphical testing has developed in statistics.

@@

One useful tool for testing statistical graphics is the concept of a lineup. 
@buja2009statistical introduced the lineup protocol in which data plots are depicted and interpreted as statistics. 
Supported by the grammar of graphics, a data plot can be characterized as a statistic, defined as, “a functional mapping of a variable or set of variables” [@vanderplas2020testing]. 
This allows the data plot to be tested similar to other statistics, by comparing the actual data plot
to a set of plots with the absence of any data structure we can test the likelihood of any perceived structure being significant. 
The construction of data plots as statistics allow for easy experimentation, granting researchers the ability to compare the effectiveness of and understand the perception of different types of charts [@vanderplas2017clusters; @vanderplas2015spatial; @hofmann2012graphical]. The lineup protocol is one such example of the development of tools designed for statistical graphical testing. The advancement of graphing software provides the tools necessary to develop new methods of testing graphics.

@@

In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing. 
The 'You Draw It' method is validated by replicating the study conducted by @mosteller1981eye. 
Based on previous research surrounding "ensemble perception," we hypothesize that regression lines fit by human perception resemble regression lines based on the principal axis rather than an ordinary least squares regression line.

@@

Participants were recruited through through Twitter, Reddit, and direct email in May 2021. 
A total of 35 individuals completed 119 unique 'You Draw It' task plots; all completed you draw it task plots were included in the analysis. 
All participants had normal or corrected to normal vision and signed an informed consent form.
The experimental tasks took approximately 15 minutes to complete.
Participants completed the experiment on their own computers in an environment of their choosing. The experiment was conducted and distributed through an RShiny application found [here](https://shiny.srvanderplas.com/you-draw-it/).

@@

The intent of research was to adapt 'You Draw It' from the New York Times feature as a tool and method for testing graphics
The study conducted by @mosteller1981eye was replicated using the 'You Draw It' method providing support for the validity of the tool.
Our results indicated that when shown points following a linear trend, participants tended to fit a line closer to a regression based on the principal axis over the ordinary least squares regression line.
This trend was most prominent when shown data simulated with larger variances. 
These results indicate that humans perform "ensemble perception" in a statistical graphic setting as participants minimized the distance from the their regression line over both the $x$ and $y$ axis simultaneously. 
In addition, we allowed participants to draw trend lines that deviated from a straight line providing insight into the curvature the human eye perceives in a set of points.

@@

This study provided a basis for the use of 'You Draw It' as a tool for testing graphics as well as provided support for "ensemble perception" in statistical graphics.
Further investigation is necessary to implement the method in non-linear settings and with real data.
This tool could also be used to evaluate human ability to extrapolate data from trends. 
In the future, an R package designed for easy implementation of 'You Draw It' task plots will help make this tool accessible to other researchers.

>>>>

authors: 
- name: Emily A. Robinson 1
  # thanks: The authors gratefully acknowledge ...
  affiliation: Department of Statistics, University of Nebraska - Lincoln

@@

abstract: |
  How do statistical regression results compare to intuitive, visually fitted results? Fitting lines by eye through a set of points has been explored since the 20th century. Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. In 2015, the New York Times introduced an interactive feature, called 'You Draw It', where readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality. In this paper, we validate ‘You Draw It’ as a method for graphical testing, comparing results to the less technological method utilized in \citet{mosteller1981eye} and extending that study with formal statistical analysis methods. Results were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least-squares regression line. This trend was most prominent when shown data simulated with larger variances. This study reinforces the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.

@@

# Introduction
We all use statistical graphics, but how do we know that the graphics we use are communicating properly? When creating a graphic, we must consider the design choices most effective for conveying the intended result. For instance, we must decide to highlight the relationship between two variables in a scatterplot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
These design choices require that we understand the perception and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.

Much of the research on the perception of visual features in charts has been conducted in psychophysics and test for accuracy and quantitative comparisons when understanding a plot. 
Cleveland and McGill conducted a series of cognitive tasks designed to establish a hierarchy of visual components for making comparisons [@cleveland1984graphical]. For example, it is more effective to display information on an $x$ or $y$ axis rather than using color in order to reduce the visual effort necessary to make numerical comparisons.
Other early studies evaluated human perception of lines. 
For instance, @cleveland1985graphical found that assessing the position of points along an axis is easier than determining the slope of a line.
The results of these cognitive tasks provide some consistent guidance for chart design; however, other methods of visual testing can further evaluate design choices and help us understand cognitive biases related to the evaluation of statistical charts.

@@

Graphical tests are useful for studying the perception of statistical graphs. 
Studies might ask participants to identify differences in graphs, read information off of a chart accurately, use data to make correct real-world decisions, or predict the next few observations.
All of these types of tests require different levels of use and manipulation of the information being presented in the chart. 
Early researchers studied graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception]. 
These studies generally tested participants ability to detect a stimulus or a difference between two stimuli.
In statistical graphics research, @vanderplas2015signs uses methods of psychophysics to estimate the effect of sign illusions by adjusting the sine illusion up and down.

@@

Efforts in the field of statistical graphics have developed graphical testing tools and methods such as the lineup protocol [@buja2009statistical] to provide a framework for inferential testing.
Through experimentation, methods such as the lineup protocol allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation [@vanderplas2017clusters; @vanderplas2015spatial; @hofmann2012graphical].
One useful tool for testing statistical graphics is the concept of a lineup.
@buja2009statistical introduced the lineup protocol in which data plots are depicted and interpreted as statistics.
Supported by the grammar of graphics, a data plot can be characterized as a statistic, defined as, “a functional mapping of a variable or set of variables” [@vanderplas2020testing].
This allows the data plot to be tested similar to other statistics, by comparing the actual data plot
to a set of plots with the absence of any data structure we can test the likelihood of any perceived structure being significant.
The construction of data plots as statistics allow for easy experimentation, granting researchers the ability to compare the effectiveness of and understand the perception of different types of charts [@vanderplas2017clusters; @vanderplas2015spatial; @hofmann2012graphical].
The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.

@@

In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing. 
The 'You Draw It' method is validated by replicating the study conducted by @mosteller1981eye. 
Based on previous research surrounding "ensemble perception," we hypothesize that visual regression tends to mimic principle component regression rather than a ordinary least squares regression.
In order to assess this hypothesis, we introduce a method for statistically modeling the participant drawn lines using generalized additive mixed models.

@@

Participants were recruited through through Twitter, Reddit, and direct email in May 2021.
A total of 35 individuals completed 119 unique 'You Draw It' task plots; all completed you draw it task plots were included in the analysis. 
All participants had normal or corrected to normal vision and signed an informed consent form.
The experimental tasks took approximately 15 minutes to complete.
While convenience sampling took place, this is primarily a perceptual task and previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
Data collected in this study serves the purpose of a pilot study intended to lay the foundation for future data collection and provides promising results for understanding the perception of regression.
Participants completed the experiment on their own computers in an environment of their choosing. The experiment was conducted and distributed through an RShiny application found [here](https://shiny.srvanderplas.com/you-draw-it/).

@@

The intent of research was to adapt 'You Draw It' from the New York Times feature as a tool and method for testing graphics and introduce a method for statistically modeling the participant drawn lines.
We provided support for the validity of the 'You Draw It' method by replicating the study found in @mosteller1981eye.
Using generalized additive mixed models, we assessed the deviation of the participant drawn lines from the statistically fitted regression lines.
Our results found that when shown points following a linear trend, participants visually fit a regression line that mimics the first principle component regression as opposed to ordinary least squares regression. 
Data simulated with a larger variance provided strong support for a participants tendency to visually fit the first principle component regression.
Our results indicate that participants minimized the distance from the their regression line over both the $x$ and $y$ axis simultaneously. 
These results provide support that humans perform "ensemble perception" in a statistical graphic setting.
We allowed participants to draw trend lines that deviated from a straight line and gained an insight into the curvature the human eye perceives in a set of points.

@@

This study provided a basis for the use of 'You Draw It' as a tool for testing statistical graphics and introduced a method for statistically modeling participant drawn lines using generalized additive mixed models.
Further investigation is necessary to implement this method in non-linear settings and with real data.
This tool could also be used to evaluate human ability to extrapolate data from trends. 
In the future, an R package designed for easy implementation of 'You Draw It' task plots will help make this tool accessible to other researchers.

\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

\begin{description}

\item[Participant Data:] De-identified participant data can be found at ...

\item[Data Analysis Code:]

\end{description}

### 2022-01-24 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
working on submission

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

% Pandoc citation processing

@@

  \author{
        Emily A. Robinson 1 \\
    Department of Statistics, University of Nebraska - Lincoln\\
     and \\     Susan VanderPlas 2 \\
    Department of Statistics, University of Nebraska - Lincoln\\
     and \\     Reka Howard 3 \\
    Department of Statistics, University of Nebraska - Lincoln\\
      }
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
Fitting lines by eye through a set of points has been explored since the
20th century. Common methods of fitting trends by eye involve
maneuvering a string, black thread, or ruler until the fit is suitable,
then drawing the line through the set of points. In 2015, the New York
Times introduced an interactive feature, called `You Draw It'. Readers
are asked to input their own assumptions about various metrics and
compare how these assumptions relate to reality. The New York Times team
utilizes Data Driven Documents (D3) that allows readers to predict these
metrics by drawing a line on their computer screen with their computer
mouse. In my research, I established `You Draw It' as a method for
graphical testing by adapting the New York Times feature. I recruited
participants via crowdsourcing websites and replicated the study found
in Eye Fitting Straight Lines (Mosteller et al., 1981). Participants
were directed to an RShiny application link and shown points following a
linear trend and asked to draw a line through the data points using
their computer mouse; task plots were generated using the r2d3 package
in R statistical software. Results from my study were consistent with
those found in the previous study; when shown points following a linear
trend, participants tended to fit the slope of the first principal
component over the slope of the least-squares regression line. This
trend was most prominent when shown data simulated with larger
variances. The reproducibility of these results serves as evidence of
the reliability of the you draw it method. Future work is necessary to
implement the `You Draw It' tool as a method of testing graphics. {[}200
word limit{]}
\end{abstract}

@@

@@

Data visualization has become a central tool in modern data science.
\citet{unwin2020data} defines data visualization as the art of graphic
displays to show data. Graphics illustrate the relationship between two
sets of information or variables. Graphical displays are useful for data
cleaning, exploratory data analysis, identifying trends, evaluating
model output, and presenting results. When creating a graphic, we must
consider the design choices most effective for conveying the intended
result. In order to evaluate the effectiveness of a graphic, we must
understand cognitive perception and biases related to different types of
data structures. Efforts in the field of graphics have developed
graphical testing tools and methods such as the lineup protocol
\citep{buja2009statistical} to provide a framework for inferential
testing. Through experimentation, methods such as the lineup protocol
allow researchers to conduct studies geared at understanding human
ability to conduct tasks related to the perception of statistical charts
such as differentiation, prediction, estimation, and extrapolation.

\hypertarget{graph-perception}{%
\subsection{Graph Perception}\label{graph-perception}}

With the widespread use of graphics, work has been done to understand
the effect of charts and graphs on human perception and higher cognitive
processing such as learning and analysis \citep{green2009personal}.
\citet{carpenter1998model} established the process of graph
comprehension in which viewers interact with charts by first
perceptually observing the visual features and later translating to
cognitive processing of the information depicted by those features. Much
of the research on the perception of visual features in charts has been
conducted in psychophysics and test for accuracy and quantitative
comparisons when understanding a plot. Cleveland and McGill conducted a
series of cognitive tasks designed to establish a hierarchy of visual
components for making comparisons \citep{cleveland1984graphical}. For
example, it is more effective to display information on an \(x\) or
\(y\) axis rather than using color in order to reduce the visual effort
necessary to make numerical comparisons. Other early studies evaluated
human perception of lines. For instance, \citet{cleveland1985graphical}
found that assessing the position of points along an axis is easier than
determining the slope of a line. The results of these cognitive tasks
provide some consistent guidance for chart design; however, other
methods of visual testing can further evaluate design choices and help
us understand cognitive biases related to the evaluation of statistical
charts.

@@

Graphical tests are useful for studying the perception of statistical
graphs. Studies might ask participants to identify differences in
graphs, read information off of a chart accurately, use data to make
correct real-world decisions, or predict the next few observations. All
of these types of tests require different levels of use and manipulation
of the information being presented in the chart. Early researchers
studied graphs from a psychological perspective
\citep{spence1990visual, lewandowsky1989perception}. These studies
generally tested participants ability to detect a stimulus or a
difference between two stimuli. Here we focus on the how graphical
testing has developed in statistics.

@@

One useful tool for testing statistical graphics is the concept of a
lineup. \citet{buja2009statistical} introduced the lineup protocol in
which data plots are depicted and interpreted as statistics. Supported
by the grammar of graphics, a data plot can be characterized as a
statistic, defined as, ``a functional mapping of a variable or set of
variables'' \citep{vanderplas2020testing}. This allows the data plot to
be tested similar to other statistics, by comparing the actual data plot
to a set of plots with the absence of any data structure we can test the
likelihood of any perceived structure being significant. The
construction of data plots as statistics allow for easy experimentation,
granting researchers the ability to compare the effectiveness of and
understand the perception of different types of charts
\citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
The lineup protocol is one such example of the development of tools
designed for statistical graphical testing. The advancement of graphing
software provides the tools necessary to develop new methods of testing
graphics.

@@

In this paper, we establish `You Draw It', adapted from the New York
Times feature, as a tool for graphical testing. The `You Draw It' method
is validated by replicating the study conducted by
\citet{mosteller1981eye}. Based on previous research surrounding
``ensemble perception,'' we hypothesize that regression lines fit by
human perception resemble regression lines based on the principal axis
rather than an ordinary least squares regression line.

@@

Participants were recruited through through Twitter, Reddit, and direct
email in May 2021. A total of 35 individuals completed 119 unique `You
Draw It' task plots; all completed you draw it task plots were included
in the analysis. All participants had normal or corrected to normal
vision and signed an informed consent form. The experimental tasks took
approximately 15 minutes to complete. Participants completed the
experiment on their own computers in an environment of their choosing.
The experiment was conducted and distributed through an RShiny
application found
\href{https://shiny.srvanderplas.com/you-draw-it/}{here}.

@@

The intent of research was to adapt `You Draw It' from the New York
Times feature as a tool and method for testing graphics The study
conducted by \citet{mosteller1981eye} was replicated using the `You Draw
It' method providing support for the validity of the tool. Our results
indicated that when shown points following a linear trend, participants
tended to fit a line closer to a regression based on the principal axis
over the ordinary least squares regression line. This trend was most
prominent when shown data simulated with larger variances. These results
indicate that humans perform ``ensemble perception'' in a statistical
graphic setting as participants minimized the distance from the their
regression line over both the \(x\) and \(y\) axis simultaneously. In
addition, we allowed participants to draw trend lines that deviated from
a straight line providing insight into the curvature the human eye
perceives in a set of points.

@@

This study provided a basis for the use of `You Draw It' as a tool for
testing graphics as well as provided support for ``ensemble perception''
in statistical graphics. Further investigation is necessary to implement
the method in non-linear settings and with real data. This tool could
also be used to evaluate human ability to extrapolate data from trends.
In the future, an R package designed for easy implementation of `You
Draw It' task plots will help make this tool accessible to other
researchers.

>>>>

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}

@@

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

  \author{
        Emily A. Robinson 1 \thanks{The authors gratefully acknowledge
\ldots{}} \\
    Department of Statistics, University of Nebraska - Lincoln\\
     and \\     Susan VanderPlas 2 \\
    Department of Statistics, University of Nebraska - Lincoln\\
     and \\     Reka Howard 3 \\
    Department of Statistics, University of Nebraska - Lincoln\\
      }
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
How do statistical regression results compare to intuitive, visually
fitted results? Fitting lines by eye through a set of points has been
explored since the 20th century. Common methods of fitting trends by eye
involve maneuvering a string, black thread, or ruler until the fit is
suitable, then drawing the line through the set of points. In 2015, the
New York Times introduced an interactive feature, called `You Draw It',
where readers are asked to input their own assumptions about various
metrics and compare how these assumptions relate to reality. In this
paper, we validate `You Draw It' as a method for graphical testing,
comparing results to the less technological method utilized in
\citet{mosteller1981eye} and extending that study with formal
statistical analysis methods. Results were consistent with those found
in the previous study; when shown points following a linear trend,
participants tended to fit the slope of the first principal component
over the slope of the least-squares regression line. This trend was most
prominent when shown data simulated with larger variances. This study
reinforces the differences between intuitive visual model fitting and
statistical model fitting, providing information about human perception
as it relates to the use of statistical graphics.
\end{abstract}

@@

--\textgreater{}

@@

We all use statistical graphics, but how do we know that the graphics we
use are communicating properly? When creating a graphic, we must
consider the design choices most effective for conveying the intended
result. For instance, we must decide to highlight the relationship
between two variables in a scatterplot by including a trend line, or
adding color to highlight clustering \citep{vanderplas2017clusters}.
These design choices require that we understand the perception and
visual biases that come into play when creating graphics, and as
graphics are evaluated visually, we must use human testing to ground our
understanding in empiricism.

Much of the research on the perception of visual features in charts has
been conducted in psychophysics and test for accuracy and quantitative
comparisons when understanding a plot. Cleveland and McGill conducted a
series of cognitive tasks designed to establish a hierarchy of visual
components for making comparisons \citep{cleveland1984graphical}. For
example, it is more effective to display information on an \(x\) or
\(y\) axis rather than using color in order to reduce the visual effort
necessary to make numerical comparisons. Other early studies evaluated
human perception of lines. For instance, \citet{cleveland1985graphical}
found that assessing the position of points along an axis is easier than
determining the slope of a line. The results of these cognitive tasks
provide some consistent guidance for chart design; however, other
methods of visual testing can further evaluate design choices and help
us understand cognitive biases related to the evaluation of statistical
charts.

@@

Graphical tests are useful for studying the perception of statistical
graphs. Studies might ask participants to identify differences in
graphs, read information off of a chart accurately, use data to make
correct real-world decisions, or predict the next few observations. All
of these types of tests require different levels of use and manipulation
of the information being presented in the chart. Early researchers
studied graphs from a psychological perspective
\citep{spence1990visual, lewandowsky1989perception}. These studies
generally tested participants ability to detect a stimulus or a
difference between two stimuli. In statistical graphics research,
\citet{vanderplas2015signs} uses methods of psychophysics to estimate
the effect of sign illusions by adjusting the sine illusion up and down.

@@

Efforts in the field of statistical graphics have developed graphical
testing tools and methods such as the lineup protocol
\citep{buja2009statistical} to provide a framework for inferential
testing. Through experimentation, methods such as the lineup protocol
allow researchers to conduct studies geared at understanding human
ability to conduct tasks related to the perception of statistical charts
such as differentiation, prediction, estimation, and extrapolation
\citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
One useful tool for testing statistical graphics is the concept of a
lineup. \citet{buja2009statistical} introduced the lineup protocol in
which data plots are depicted and interpreted as statistics. Supported
by the grammar of graphics, a data plot can be characterized as a
statistic, defined as, ``a functional mapping of a variable or set of
variables'' \citep{vanderplas2020testing}. This allows the data plot to
be tested similar to other statistics, by comparing the actual data plot
to a set of plots with the absence of any data structure we can test the
likelihood of any perceived structure being significant. The
construction of data plots as statistics allow for easy experimentation,
granting researchers the ability to compare the effectiveness of and
understand the perception of different types of charts
\citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
The advancement of graphing software provides the tools necessary to
develop new methods of testing statistical graphics.

@@

In this paper, we establish `You Draw It', adapted from the New York
Times feature, as a tool for graphical testing. The `You Draw It' method
is validated by replicating the study conducted by
\citet{mosteller1981eye}. Based on previous research surrounding
``ensemble perception,'' we hypothesize that visual regression tends to
mimic principle component regression rather than a ordinary least
squares regression. In order to assess this hypothesis, we introduce a
method for statistically modeling the participant drawn lines using
generalized additive mixed models.

@@

Participants were recruited through through Twitter, Reddit, and direct
email in May 2021. A total of 35 individuals completed 119 unique `You
Draw It' task plots; all completed you draw it task plots were included
in the analysis. All participants had normal or corrected to normal
vision and signed an informed consent form. The experimental tasks took
approximately 15 minutes to complete. While convenience sampling took
place, this is primarily a perceptual task and previous results have
found few differences between expert and non-expert participants in this
context \citep{vanderplas2015spatial}. Data collected in this study
serves the purpose of a pilot study intended to lay the foundation for
future data collection and provides promising results for understanding
the perception of regression. Participants completed the experiment on
their own computers in an environment of their choosing. The experiment
was conducted and distributed through an RShiny application found
\href{https://shiny.srvanderplas.com/you-draw-it/}{here}.

@@

The intent of research was to adapt `You Draw It' from the New York
Times feature as a tool and method for testing graphics and introduce a
method for statistically modeling the participant drawn lines. We
provided support for the validity of the `You Draw It' method by
replicating the study found in \citet{mosteller1981eye}. Using
generalized additive mixed models, we assessed the deviation of the
participant drawn lines from the statistically fitted regression lines.
Our results found that when shown points following a linear trend,
participants visually fit a regression line that mimics the first
principle component regression as opposed to ordinary least squares
regression. Data simulated with a larger variance provided strong
support for a participants tendency to visually fit the first principle
component regression. Our results indicate that participants minimized
the distance from the their regression line over both the \(x\) and
\(y\) axis simultaneously. These results provide support that humans
perform ``ensemble perception'' in a statistical graphic setting. We
allowed participants to draw trend lines that deviated from a straight
line and gained an insight into the curvature the human eye perceives in
a set of points.

@@

This study provided a basis for the use of `You Draw It' as a tool for
testing statistical graphics and introduced a method for statistically
modeling participant drawn lines using generalized additive mixed
models. Further investigation is necessary to implement this method in
non-linear settings and with real data. This tool could also be used to
evaluate human ability to extrapolate data from trends. In the future,
an R package designed for easy implementation of `You Draw It' task
plots will help make this tool accessible to other researchers.

\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

\begin{description}

\item[Title:] Brief description. (file type)

\item[R-package for  MYNEW routine:] R-package <U+FFFD>MYNEW<U+FFFD> containing code to perform the diagnostic methods described in the article. The package also contains all datasets used as examples in the article. (GNU zipped tar file)

\item[HIV data set:] Data set used in the illustration of MYNEW method in Section~ 3.2. (.txt file)

\end{description}
